\pkg{shorts}: An R Package for Modeling Short Sprints

### 2020-10-21 paper.Rmd
Draft ready

---
documentclass: jss
author:
  - name: Mladen Jovanović
    affiliation: |
      | University of Belgrade
    address: |
      | Faculty of Sport and Physical Education
      | Belgrade, Serbia
    email: \email{coach.mladen.jovanovic@gmail.com}
  - name: Jason D. Vescovi
    affiliation: |
      | University of Toronto
    address: |
      | Faculty of Kinesiology and Physical Education
      | Graduate School of Exercise Science
      | Toronto, ON Canada
    email: \email{vescovij@gmail.com}
title:
  formatted: "\\pkg{shorts}: An R Package for Modeling Short Sprints"
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). Short sprint performance data comes from two sources: (1) split times using timing gates and (2) radar gun. In this paper, we demonstrate how these data can be modeled using the \pkg{shorts} package and how timing issues (i.e., synchronization issues with the radar gun, or using the flying start with the timing gates), which are common in practical measurement settings, can introduce error in parameter estimates. Further, we demonstrated how different model definitions implemented in \pkg{shorts} package can alleviate these errors.
keywords:
  formatted: [sprints, sports, modeling]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
citation_package: natbib
bibliography: [references.bib, packages.bib]
biblio-style: apalike
link-citations: yes
editor_options: 
  chunk_output_type: console
---

@@

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
), "packages.bib")

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 sec) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

>>>>

---
documentclass: jss
author:
  - name: Mladen Jovanović
    affiliation: |
      | University of Belgrade
    address: |
      | Faculty of Sport and Physical Education
      | Belgrade, Serbia
    email: \email{coach.mladen.jovanovic@gmail.com}
  - name: Jason D. Vescovi
    affiliation: |
      | University of Toronto
    address: |
      | Faculty of Kinesiology and Physical Education
      | Graduate School of Exercise Science
      | Toronto, ON Canada
    email: \email{vescovij@gmail.com}
title:
  formatted: "\\pkg{shorts}: An R Package for Modeling Short Sprints"
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \pkg{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \pkg{shorts} package to help alleviate errors within the subsequent parameter outcomes.
keywords:
  formatted: [sprinting, sports, modeling]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
citation_package: natbib
bibliography: [references.bib, packages.bib]
link-citations: yes
editor_options: 
  chunk_output_type: console
---

@@

# automatically create a bib database for R packages
#knitr::write_bib(c(
#  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW"
#), "packages.bib")

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

### 2020-10-21 paper.tex
Draft ready

\Abstract{
Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). Short sprint performance data comes from two sources: (1) split times using timing gates and (2) radar gun. In this paper, we demonstrate how these data can be modeled using the \pkg{shorts} package and how timing issues (i.e., synchronization issues with the radar gun, or using the flying start with the timing gates), which are common in practical measurement settings, can introduce error in parameter estimates. Further, we demonstrated how different model definitions implemented in \pkg{shorts} package can alleviate these errors.
}

\Keywords{sprints, sports, modeling}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 sec) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

>>>>

\Abstract{
Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the \pkg{shorts} package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the \pkg{shorts} package to help alleviate errors within the subsequent parameter outcomes.
}

\Keywords{sprinting, sports, modeling}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5s) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.
